Estimating Predictive Rate-Distortion via Neural Variational Inference

### 2019-03-05 paper.tex

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

@@

\section{Theoretical Discussion}

\subsection{A Family of Examples}
We present a family of processes that have finite excess entropy and infinite statistical complexity, and which may provide a simple mathematical model of some aspects of language structure.

\begin{defin}
Let $\mu$ be any distribution on the nonnegative integers such that $\mu(0) > 0$.
Then consider the process $(X_t)_t$ where the marginal of $X_t$ is uniform on $\{0,1\}$, and where $X_t$ is generated by (i) taking an independent sample $k \sim \mu$, and (ii) setting $X_t := X_{t-k}$ if $k > 0$, and $X_t$ a fresh uniform draw from $\{0,1\}$ if $k=0$.
\end{defin}

\begin{remark}
I don't know to what extent it is `obvious' that such a process actually exists. I think one can show it by taking the product of the process $K_i \sim_{iid} \mu$, which leads to a partitioning of $\mathbb{Z}$ into equivalence classes of positions that must have the same value in $(X_i)_i$, together with an iid p=1/2 Bernoulli process, which supplies the values to these equivalence classes. Formally, $(X_i)_i$ is then a measurable function of a product of the $K_i$ process and a Bernoulli process, and thus exists.
By marginalizing out $K_i$ given $X_t$, one can verify that this process has the property used to define it in the definition.
\end{remark}

%and the iid p=1/2 Bernoulli process, and condition on $X_$

%\textbf{TODO} is it clear that this process actually exists? Can we invoke Kolmogorov extension thm? Or is this trivial? In any case, we can approximate it by n-th order Markov processes whose excess entropies stay bounded whereas their statistical complexities go to infinity (and which converge to the intended process in average KL distance) (?) In this sense, it may not be a disaster if the process turned out not to exist (?)

Let us first consider how such a process may relate to natural language.

TODO talk a bit about dependencies, agreement, ...

We can prove that it has finite excess entropy and infinite complexity under very general conditions:

%Now examine its statistical complexity and excess entropy:

\begin{proposition}
The following two statements hold:

If $\mu$ has infinite support (i.e., $\#\{k : \mu(\{k\}) > 0\} = \infty$), then $(X_t)_t$ has infinite statistical complexity.

If $\E_{k\sim \mu}[k^2] < \infty$, then $(X_t)_t$ has finite excess entropy.
\end{proposition}
There are many measures $\mu$ that satisfy both constraints, e.g. $\mu(x) \propto 2^{-x}$ or $\mu(x) \propto x^{-4}$.

\begin{proof}
We first show that the process has finite excess entropy:
\begin{align*}
I[X_{<0}, X_{>0}] &= \sum_{t=1}^\infty t I[X_t, X_0|X_1\dots X_{t-1}] \\
& \leq \sum_{t=1}^\infty t \operatorname{\mu}(\{k : k \geq t\}) \\
&= \sum_{t=1}^\infty \frac{t(t-1)}{2} \operatorname{\mu}(\{t\}) \\
&= \frac{1}{2} \E_{k \sim \mu} \left[k^2-k\right]
\end{align*}
which is finite by assumption.

We now show that the process has infinite statistical complexity.

%In the special case where $\mu(n) = 2^{-n}$, it is clear that the causal states are the semi-infinite pasts.
%In general, I don't know whether that's true, but we can still prove infinite complexity.

%and no loss-less compression of pasts is possible. To see this, consider the probabilities $$P\left(X_T=1|X_1=0,...,X_{T-1}=0; x_{<0}\right) = 1/2 \mu(0) + \sum_{t=T}^\infty 1_{x_{T-t} = 1} \mu(t)$$TODO and then?

Since $\mu$ has infinite support and integrates to a finite value, we can find an infinite ascending sequence $0 < t_1 < t_2 < ...$ of integers such that
$$\mu\left(t_i < K \leq t_{i+1}\right) > \mu(K > t_{i+1})$$
for all integers $i$.
Now consider the set $\mathcal{X} \subset \{0,1\}^{\mathbb{N}}$ of semi-infinite sequences that are constant (either 0 or 1) on each interval $t_i, ..., t_{(i+1)}-1$.

We will show that the elements of $\mathcal{X}$ represent distinct causal states.
%Recall that the statistical complexity is defined as the entropy of the causal states.
%(Note that, as $\mathcal{X}$ is uncountable and each of its elements has measure zero, some care has to be taken here.)
%We will show that all the prefixes $x_{1\dots t_i}$ ($i \in \mathbb{N}$), for $x_{1,2,...} \in \mathcal{X}$, 
%$$P\left(X_{t_j}=1|x_{1...t_j-1}; x_{<0}\right) = 1/2 \mu(0) + \sum_{t=t_j}^{t_{j+1}-1} 1_{x_{t} = 1} \mu(t)$$
That is, we show that one can uniquely identify elements of $\mathcal{X}$ from the induced distributions over futures.
Take a semi-infinite sequence $x_{1...} \in \mathcal{X}$.
Assume the distribution $P\left(X_{-1}=1|x_{1}...\right)$ is known, and $x_{1}...x_{t_i-1}$ have already been identified from it.
Then 
$$P\left(X_{-1}=1|x_{1}...\right) = 1/2 \mu(0) + \sum_{t=1}^{t_{i}-1} 1_{x_{t} = 1} \mu(t) + \sum_{t=t_i}^\infty 1_{x_{t} = 1} \mu(t)$$
where $1/2 \mu(0) + \sum_{t=1}^{t_{i}-1} 1_{x_{t} = 1} \mu(t)$ is already a known quantity.
For the remaining part $C_i := \sum_{t=t_i}^\infty 1_{x_{t} = 1} \mu(t)$, we have
$$C = \sum_{t=t_i}^\infty 1_{x_{t} = 1} \mu(t) = \sum_{t=t_i}^{t_{i+1}-1} 1_{x_{t} = 1} \mu(t) + \sum_{t=t_{i+1}}^\infty 1_{x_{t} = 1} \mu(t)$$
which is equal to 
$$\mu(\{t_i...t{i+1}-1\}) + \sum_{t=t_{i+1}}^\infty 1_{x_{t} = 1} \mu(t)$$
or
$$\sum_{t=t_{i+1}}^\infty 1_{x_{t} = 1} \mu(t)$$
depending on whether $X$ is $1$ or $0$ on the set $\{t_i...t_{i+1}-1\}$. Given that 
$$\sum_{t=t_{i+1}}^\infty 1_{x_{t} = 1} \mu(t) \leq \mu(K \geq t_{i+1}) < \mu(\{t_i...t{i+1}-1\})$$
by assumption, the block $x_{t_i...t{i+1}-1}$ can be identified from $C_i$.
%TODO rewrite this
%- first identify the first block
%- then shift, can compute remainder
%Now for any $t_j$ consider the probability
%$$P\left(X_{t_j}=1|X_1=0,...,X_{t_j-1}=0; x_{-1}...x_{t_i-t_j}\right) = 1/2 \mu(0) + \sum_{t=t_j}^{t_{j+1}-1} 1_{x_{t} = 1} \mu(t) + \sum_{t=t_{j+1}}^{\infty} 1_{x_{t} = 1} \mu(t)$$
%where $x_{<0} \in \mathcal{X}$.
%By the previous inequality, this probability is sufficient to identify whether $X_{t_{j-1}}...X_{t_j-1}$ is identically 0 or 1.
Thus, the elements of $\mathcal{X}$ yield different distributions over future observations, and thus represent distinct causal states of the process (modulo the complication that each of them has measure zero).

Indeed, we have shown that the prefixes $x_{1\dots t_i}$ ($i \in \mathbb{N}$), for $x_{1,2,...} \in \mathcal{X}$, represent disjoint sets of causal sets.
For all integers $i$ and any sequence $X_1^\infty \in \mathcal{X}$, the block $x_{1...t_i}$ has log-probability at least $t_i \log \frac{\mu(0)}{2} = \Theta(t_i)$.
Thus
$$H[X_1...X_{t_i} | X \in \mathcal{X}] \geq  t_i \log \frac{\mu(0)}{2} = \Theta(t_i)$$
In conclusion, the statistical complexity is infinite.

%For any positive integer $T$ and any string $x_1...x_T \in \{0,1\}^T$, we have $\log P(X_1...X_T=x_1...x_T) \geq T \log\frac{\mu(\{0\})}{2}$, showing that $$H[X_1...X_T] \geq T \log\frac{\mu(\{0\})}{2} \rightarrow_{T \rightarrow\infty}\infty$$

\end{proof}

\paragraph{Bounding the PRD Curve}
We can bound the predictive rate-distortion curve of this process as follows:
Using an $n$-word context incurs rate at least $n \log \frac{\mu(0)}{2}$, and distortion at most $\sum_{k > n} (k-n) \mu(k > n)$. That is, the faster $\mu(t)$ goes to zero as $t \rightarrow \infty$, the more favorable the prediction-rate tradeoff is (as far as quantified by this upper bound).
%Matching with our results for natural language would suggest $\mu(k>n) \sim n^{1/\beta-1} \exp(-\Theta(n^{1/\beta}))$ -- that is, we expect the probability of long dependencies (within sentences) to decay very fast.

We consider a code $Z_T$ that stores the last $T$ bits.
The rate of this code is $\Theta(T)$ (since $\mu(0) > 0$).
(This code might not be optimal.)

Its distortion is $I[X_{\leq 0}, X_{>T}|X_{1\dots T}]$.
This is upper-bounded by the probability that any future $K$ would reach back to a bit earlier than the last $T$ bits.\footnote{
\begin{align*}
I[X_{\leq 0}, X_{>T}|X_{1\dots T}] & = \sum_{t=T}^\infty (t-T) I[X_t, X_0|X_1\dots X_{t-1}] \\
& \leq \sum_{t=T}^\infty (t-T) \operatorname{\mu}(\{k : k \geq t\}) \\
& = \sum_{t=1}^\infty t \operatorname{\mu}(\{k : k \geq t+T\}) \\
&= \sum_{t=1}^\infty \frac{t(t-1)}{2} \operatorname{\mu}(\{t+T\}) \\
&= \sum_{t=T}^\infty \frac{(t-T)(t-1-T)}{2} \operatorname{\mu}(\{t\}) \\
&= 1/2 \cdot \E[(k-T)_+^2-(k-T)_+]
\end{align*}}
and a lower bound can similarly be derived. \footnote{$I[X_t, X_0|X_1\dots X_{t-1}]$ is at least the probability that the block did not copy from $X_0$, but $X_t$ does.
This is $(1-\mu(1))\dots(1-\mu(t-1))\mu(t)$. So we get
\begin{align*}
I[X_{\leq 0}, X_{>T}|X_{1\dots T}] & = \sum_{t=T}^\infty (t-T) I[X_t, X_0|X_1\dots X_{t-1}] \\
& \geq \sum_{t=T}^\infty (t-T) (1-\mu(1))\dots(1-\mu(t-1))\mu(t) \\
& \geq \left(\sum_{t=T}^\infty (t-T)\mu(t) \right) \left( \prod_{j=1}^\infty (1-\mu(j))\right) \\
& = \E \left[(t-T)_+ \right] \left( \prod_{j=1}^\infty (1-\mu(j))\right) \\
\end{align*}}
We get that the distortion of the code $Z_T$ is bounded as 
\begin{equation}
\E \left[(t-T)_+ \right] \left( \prod_{j=1}^\infty (1-\mu(j))\right) \leq I[X_{\leq 0}, X_{>T}|X_{1\dots T}] \leq 1/2 \cdot \E[(k-T)_+^2-(k-T)_+]
\end{equation}

If $\mu(k) = q^{k} \cdot (1-q)$ (which is a probability distribution over $\mathbb{N}$ for $q < 1$), we have(TODO probably there are algebra errors here)\footnote{$\E \left[(t-T)_+ \right] = \E[k] \mu(k\geq T)$
$\E \left[(t-T)_+^2 \right] = \E[k^2] \mu(k\geq T) = \frac{q+q^2}{(1-q)^2} T q^T (1-q)$, 
$\mu(k\geq T) = T q^T (1-q)$
. where $\phi(q)$ is the Euler function, equal to the Pochhammer symbol $(q;q)_\infty$.}
\begin{equation}
T \frac{q^{T+1}}{1-q}  \phi(q) (1-q) \leq I[X_{\leq 0}, X_{>T}|X_{1\dots T}] \leq 1/2 \cdot (\frac{q+q^2}{(1-q)^2} T q^T (1-q) - T \frac{q^{T+1}}{1-q} )
\end{equation}
and thus
\begin{equation}
I[X_{\leq 0}, X_{>T}|X_{1\dots T}] = \Theta(T q^T)
\end{equation}
In conclusion, we find that $R = \Theta(T)$, $P = \Theta(C-T q^T)$.

Now look at $\mu(k) \sim k^{-t}$ where $t > 3$.
We will take $t$ as a constant here, considering only $T$ as varying.
The infinite product is a (nonzero) constant $C(t)$.
\begin{equation}
\Theta(T^{-t+2} + T^{-t+1}) \leq I[X_{\leq 0}, X_{>T}|X_{1\dots T}] \leq \Theta(T^{-t+3} + T^{-t+2})
\end{equation}
and thus
\begin{equation}
I[X_{\leq 0}, X_{>T}|X_{1\dots T}] = \Theta(T^{-t+2})
\end{equation}
We find that $D = \Theta(R^{-t+1})$, and $\frac{\partial P}{\partial R} = \Theta( R^{-t})$.

%$\frac{\partial P}{\partial R} = \Theta(q^T + T q^T)$

These bounds illustrate some of the kinds of curves that we would expect; however, I don't know how to exactly calculate the curve (as the code $Z_T$ may not be optimal).

\paragraph{Considering Individual Sentences}
This process differs from the setting of individual sentence in that the latter setting involves `boundaries' at which future and past are independent.
We can also construct a variant with this property by taking a distribution $\mu'$ over lengths $1,2,...$, and `resetting' the process in intervals of size $\sim \mu'$, and restricting future copy operations to positions to the right of the last reset point.
This process has statistical complexity %$\Theta(\E_{L \sim \mu'}[L])$: It is clearly upper-bounded by $\E_{L \sim \mu'}[L]$, but also in 
$\Omega\left(\log \frac{\mu(0)}{2} \cdot \E_{L \sim \mu'}[L]\right)$ by the same argument as we developed in the proof of the proposition.
If $\E_{L \sim \mu'}[L] \gg \E_{K \sim \mu}[K^2]$, we obtain a process with finite complexity much larger than its excess entropy, which for all practical purposes may look like its infinite-complexity counterpart.

\paragraph{Prior Work}
Debowski has discussed both the Bernoulli Mixture and the Santa Fe processes as in certain ways  analogous to natural language.
These processes, however, have zero crypticity, so their PRD curves are straight diagonals. (Restricting them to individual sentences would change this.)

%Santa Fe process where each sentence has their own fact collection:
%- best codes just store the facts most likely to be retrieved, to the extent that they are known
%at fixed sentence length
%- predictiveness is larger for concentrated $\mu$
%- smaller for diffuse $\mu$

\subsection{Predictive Rate-Distortion for Processes with Infinite Complexity}
Here we want to show that, even for processes with infinite statistical complexity, the PRD curve provides meaningful quantification of the difficulty of prediction.

\begin{proposition}
Let $(X_t)_t$ be a process such that the entropy $H[X_t]$ is finite (e.g., if the state space is finite).

Then the PRD curve, as a function of Predictiveness, is finite on the interval from 0 to (possibly excluding) the excess entropy.
\end{proposition}

\begin{proof}
An upper bound on the curve is given by the (convex hull) of the pairs of $$\{(H[X_1...X_t], I[(X_1...X_t), X_{>t}]) : t \in \mathbb{N}\}$$
all which are finite under the assumption. 
Also, $$EE = \sup_{t \in \mathbb{N}} I[(X_1...X_t), X_{>t}]$$
So for all values of predictiveness that are strictly below the EE, the corresponding rate has a finite bound.
\end{proof}

\subsection{Sensitivity/Stability of PRD curve}
Unlike statistical complexity, the PRD curve will not differ much between processes that have similar distributions, at least for finite-order Markov processes:

%\begin{proposition}
%	Fix $\epsilon > 0$. % and $M \in \mathbb{N}$.
%	Let $(Y_t)_t$ be a process distributed according to $Q$, and assume that it is close to $X_t$ (distributed according to $P$) in KL Divergence in the following sense: For all $M \in \mathbb{N}$,
%	$$\E_{X_{-\infty}...X_0} \left[\log \frac{P(X_1\dots X_{M}|X_{-\infty}...X_0)}{Q(X_1\dots X_{M}|X_{-\infty}...X_0)}\right] < \epsilon$$
%%	$$\E_{X_{-\infty}...X_0} \left[\log \frac{P(X_1|X_{-\infty}...X_0)}{Q(X_1|X_{-\infty}...X_0)}\right] < \epsilon$$
	%$D_{KL}((X_{1...M}) || Y_{1...M} | \overleftarrow{X}) < \epsilon H[S_0]$.
%	Then the predictive rate-distortion curve of $(Y_t)_t$ is, up to $\epsilon$, a (pessimistic) upper bound on the curve of $(X_t)_t$.
%\end{proposition}

\begin{proposition}
Fix $\epsilon$ and $M \in \mathbb{N}$.

	Let $(X_t)_t \sim P$, $(Y_t)_t \sim Q$ be two $M$-th order Markov processes. Assume that
	$$\E_{X_{-\infty}...X_0} \left[\log \frac{P(X_1|X_{-\infty}...X_0)}{Q(X_1|X_{-\infty}...X_0)}\right] < \epsilon$$

	For each point $(R,P$) on the PRD of $(Y_t)_t$, the point $(R+M\epsilon, P - M \epsilon)$ is an upper bound on the PRD of $(X_t)_t$.

\begin{proof}
TODO write
%Storing $S'_0$ and predicting the future on its basis results in $\epsilon H[S_0]$ additional loss in the future.
\end{proof}

As a corollary, we can conclude that the PRD curve is a continuous function of the process $(X_t)_t$, in the following sense.
We view the PRD curve as a function defined on all of $\mathbb{R}_+$ by viewing it as a function of the rate.
Then we have:

%, as a set in $\mathbb{R}^2$. %The Hausdorff distance between two curves $c_1, c_2$ is ... %let's say a sequence of curves $c_n$ converges uniformly to a curve $c$ if 
%$$\lim_{n\rightarrow\infty} \sup_{x \in c_n} \inf_{y \in c} \|x-y\| = 0$$

\begin{corollary}
If $P, P_n$ are the distributions of $M$-th order Markov processes ($M$ fixed), and if
$$D_{KL}(P_n(\cdot|X_{<0}) || P(\cdot|X_{<0})) \rightarrow 0$$
$$D_{KL}(P(\cdot|X_{<0}) || P_n(\cdot|X_{<0})) \rightarrow 0$$
then the PRD curve of $P_n$ converges to the PRD curve of $P$ uniformly on all of $\mathbb{R}_+$.

%TODO verify this!
\end{corollary}

\begin{proof}
TODO
\end{proof}

%	Then the predictive rate-distortion curve of $(Y_t)_t$ is, up to $...\epsilon$, a (pessimistic) upper bound on the curve of $(X_t)_t$:
%	For a given level of rate, the achieved predictivity of $(X_t)_t$ at this rate is at least the predictivity of $(Y_t)_t$ at this rate minus $...\epsilon$.

Such a continuity statement is not true for statistical complexity (see the next example).

\begin{example}[Simple Example]
Here is a very simple example with $M=1$.
Let $S$ be a finite set.

Process 1, with distribution $P$, consists of IID uniform draws from $S$. . %\log |S|$. % ($\{0,1\}^\mathbb{Z}$).

Process 2, with distribution $P_n$, chooses $X_{t+1}$ to be $X_t$ with probability $1/n$, and a new independent uniform draw from $S$ otherwise.

The statistical complexity is $0$ for $P$, but $\log |S|$ for any $P_n$, independent on $n$. In contrast, the PRD curve of $P_n$ converges to the PRD curve of $P$, which is the set $\mathbb{R}_+ \times \{0\}$: For any given rate, one can only achieve zero predictiveness for a process of IID samples.
The PRD curve of $P$ converges to that curve, as maximum achievable predictiveness goes to zero as $n\rightarrow \infty$.
\end{example}

I don't know how to prove anything like this when the processes do not have uniformly bounded Markov order.

In Lohr CITE it was shown that statistical complexity is lower weak-$^*$ semicontinuous, for general stochastic processes.
I don't know whether the same holds for the PRD curve.

%\begin{proposition}
%	Fix $\epsilon_1, \epsilon_2 > 0$ and $M \in \mathbb{N}$.
%	Let $(Y_t)_t$ be a process distributed according to $Q$, and assume that it is close to $X_t$ (distributed according to $P$) in KL Divergence in the following sense:
%	$$\E_{X_{-\infty}...X_0} \left[\log \frac{P(X_1\dots X_{M}|X_{-\infty}...X_0)}{Q(X_1\dots X_{M}|X_{-\infty}...X_0)}\right] < \epsilon_1$$
%	Further, assume $I[X_{>M}, X_{<0}|X_{1...M}] < \epsilon_2$.
%	$$\E_{X_{-\infty}...X_0} \left[\log \frac{P(X_1|X_{-\infty}...X_0)}{Q(X_1|X_{-\infty}...X_0)}\right] < \epsilon$$
	%$D_{KL}((X_{1...M}) || Y_{1...M} | \overleftarrow{X}) < \epsilon H[S_0]$.
%	Then the predictive rate-distortion curve of $(Y_t)_t$ is, up to $\epsilon_1 + \epsilon_2$, a (pessimistic) upper bound on the curve of $(X_t)_t$:
%	For a given level of rate, the achieved predictivity of $(X_t)_t$ at this rate is at least the predictivity of $(Y_t)_t$ at this rate minus $\epsilon_1+\epsilon_2$.

%	TODO but the rate will also change???
%\end{proposition}

>>>>

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{thm}[theorem]{Theorem}

@@
